Certifying, incremental type checking

### 2012-06-14 doc/papers/LFMTP2012/paper.tex
paper: various

We exploit here an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
won't return ill-typed values, but arguments and results are checked
at run-time at the boundaries of defined functions. This allows to
omit formal justifications, like the fact that we need to iterate
twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
course this incurs on the performance; we will see a compensation for
it by the slicing mechanism described below in \ref{sec:repr}.

@@

\section{Designing the framework}
%% Here is my idea (2 pages)

% explicit LF
% lambda-lifting

@@

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
others, and deconstruct them by case analysis. The definition $T$ of a function always begins with a series of
$\lambda$.

@@

This language is dynamically typed: there is no static guarantee that
object $\app {\fct f} x$ produced by a function $\fct f : A_1\to A_2 =
T$ will have type $A_2$, nor that pattern matching is exhaustive. It
is also non-terminating: we can define a recursive function $\fct f :
\cst a\to \cst a = \lam x \app {\fct f} {\var x}$. Nevertheless, we
want to maintain the invariant that each time an object goes in or out
of a function $\fct f : A_1\to A_2$, it is checked against the type of
this function (resp. $A_1$ or $A_2$). You can think of interpreted
functions as mere black boxes taking and producing objects, and being
able to feed objects to other black boxes; yet the wires linking them
are labeled with types and objects circulating in them are checked
against that type.

@@

\begin{definition}
  To each function $\fct f : A = T$ in $\Sigma$ we associate a family
  of \emph{inverses} $\fcti f {\;n} : (A)^n = \pi_n(A)$ for $n\in\mathbb
  N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
  \begin{align*}
    (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
    (\app {\cst a} S)^A &= \prd x {\app{\cst a} S} A \\
    \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
    \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
    \pi^y(\app{\cst a} S) &= \var y
  \end{align*}
\end{definition}

@@

    \infer[MAtom]{
      \jf\Delta\Gamma F {\Delta'} {F'} {P}
       \and
      \jea{\Delta'}\Gamma P {\app{\cst a} S}
    }{
      \jm\Delta\Gamma F {\app{\cst a} S} {\Delta'} {F'}
    }
    % TODO commenter type ground
  \end{mathpar}

@@

>>>>

We exploit here an idea similar to run-time \emph{contracts}
\cite{wadler2009well}: there is no static guarantee that computations
won't return ill-typed values, but arguments and results are checked
at run-time at the boundaries of defined functions. This allows to
omit formal justifications, like the fact that we need to iterate
twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
course this incurs on the performance.

@@

\section{Design of the framework}
%% Here is my idea (2 pages)

% TODO explicit LF
% TODO lambda-lifting

@@

Secondly, we introduce the \CL\ language to transform our \SLF\
objects (Fig.~\ref{fig:syntax-CL}). We can construct atomic objects
out of others, and deconstruct them by case analysis. The definition
$T$ of a function always begins with a series of $\lambda$.

@@

This language is untyped, but types are dynamically checked: there is
no static guarantee that object $\app {\fct f} x$ produced by a
function $\fct f : A_1\to A_2 = T$ will have type $A_2$, nor that
pattern matching is exhaustive. It is also non-terminating: we can
define a recursive function $\fct f : \cst a\to \cst a = \lam x \app
{\fct f} {\var x}$. Nevertheless, we want to maintain the invariant
that each time an object goes in or out of a function $\fct f : A_1\to
A_2$, it is checked against the type of this function (resp. $A_1$ or
$A_2$). You can think of interpreted functions as mere black boxes
taking and producing objects, and being able to feed objects to other
black boxes; yet the wires linking them are labeled with types and
objects circulating in them are checked against that type.

@@

\begin{definition}
  To each function $\fct f : A = T$ in $\Sigma$ we associate a family
  of \emph{inverses} $\fcti f {\;n} : (A)^n = \pi_n(A)$ for $n\in\mathbb
  N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
  \begin{align*}
    (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
    (\app P)^A &= \prd x P A \\
    \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
    \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
    \pi^y(P) &= \var y
  \end{align*}
\end{definition}

@@

    \infer[MAtom]{
      \jf\Delta\Gamma F {\Delta'} {F'} {P'}
       \and
      \jea{\Delta'}\Gamma P {P'}
    }{
      \jm\Delta\Gamma F P {\Delta'} {F'}
    }
    % TODO commenter type ground
  \end{mathpar}

@@

\begin{theorem}
  If $\mr$ well-constructed, and $\commit\mr F = \mmr$, then
  $\checkout{\mmr}$ is a value.
\end{theorem}
